Gut metagenome in European women with normal, impaired and diabetic glucose control
========================================================

```{r loadlibs,echo=FALSE, warning=FALSE,message=FALSE}
library(randomForest)
library(ggplot2)
library(xtable)
library(plyr)
```


This is an attempt to recreate the figures in the paper "Gut metagenome in European women with normal, impaired and diabetic glucose control". Their methods included the training of a random forest for diabetes prediction. The data used for training was obtained from the authors.


```{r readdata,cache=TRUE}
cluster.table <- read.delim("data/ClusterTable.txt", sep="\t", header=TRUE)
strain.table <- read.delim("data/StrainTable.txt", sep="\t", header=TRUE)
supptable3 <- read.csv("data/SuppTable3.csv")
```
They focus on the Type II diabetes (T2D) and normal glucose tolerance (NGT) classes. There is another class present in the sample (impaired glucose tolerance; IGT), but I will remove these data points.

```{r subsetsupptable,cache=TRUE}
supptable3 <- subset(supptable3, Classification != "IGT")
supptable3 <- transform(supptable3, Classification=factor(Classification))
```

We need to combine the classifications with the abundances:
```{r combinedata,cache=TRUE}
cluster.table <- merge(supptable3[, c("Sample.ID", "Classification")],
                       cluster.table,
                       by.x="Sample.ID", by.y="Subjects")
strain.table <- merge(supptable3[, c("Sample.ID", "Classification")],
                      strain.table,
                      by.x="Sample.ID", by.y="Subject")

cluster.data <- cluster.table[, -1]
strain.data <- strain.table[, -1]

```

The "Statistical Analysis" section of the methods says that they only included species with an abundance of at least $10^{-5}$ in any subject.

```{r topstrains,message=FALSE}
min.abundance <- 10**-5
strain.keep.abundances <- which(apply(strain.data[, -1], 2, 
                                      function(x) any(x >= min.abundance)))

# Add one to account for the column I left out
strain.keep.abundances <- strain.keep.abundances + 1

# Top 915 by mean
strain.data <- strain.data[, c(1, strain.keep.abundances)]

```

A view of number of subjects per group:

```{r histogram,fig.width=7,fig.height=6,echo=TRUE}
p <- ggplot(cluster.table, aes(x=Classification))
p <- p + geom_bar()
p
```

Let's compute the Wilcoxon statistics between the T2D and NGT groups for all strains.

```{r wilcoxfn,echo=FALSE,results='hide'}
# Helper function for computing wilcoxon and getting the p-value column-wise from a matrix
do.wilcoxon <- function(df, classif, ...) {
  colnames(df) <- "value"
  tmpdf <- data.frame(Classification=classif, value=df)
  wilcox.test(formula=value ~ Classification, data=tmpdf, ...)$p.val
}
```

```{r wilcoxon.strains,cache=TRUE}
wilcoxon.strains.gr <- adply(.data=strain.data[, -1], 2,
                             .fun=do.wilcoxon,
                             classif=strain.data[, 1],
                             alternative="g")
colnames(wilcoxon.strains.gr) <- c("Species", "p.val")
wilcoxon.strains.gr <- transform(wilcoxon.strains.gr, 
                                 p.adj=p.adjust(p.val, method="BH"))
wilcoxon.strains.gr <- wilcoxon.strains.gr[order(wilcoxon.strains.gr$p.adj, 
                                                 decreasing=FALSE), ]
wilcoxon.strains.gr <- subset(wilcoxon.strains.gr, p.adj <= 0.05)

wilcoxon.strains.le <- adply(.data=strain.data[, -1], 2,
                             .fun=do.wilcoxon,
                             classif=strain.data[, 1],
                             alternative="l")
colnames(wilcoxon.strains.le) <- c("Species", "p.val")
wilcoxon.strains.le <- transform(wilcoxon.strains.le,
                                 p.adj=p.adjust(p.val, method="BH"))
wilcoxon.strains.le <- wilcoxon.strains.le[order(wilcoxon.strains.le$p.adj,
                                                 decreasing=FALSE), ]
wilcoxon.strains.le <- subset(wilcoxon.strains.le, p.adj <= 0.05)
```

```{r diff.strains.gr,results='asis'}
tbl.strains.gr <- xtable(wilcoxon.strains.gr)
print(tbl.strains.gr, type="html")
```

```{r diff.strains.le,results='asis'}
tbl.strains.le <- xtable(wilcoxon.strains.le)
print(tbl.strains.le, type="html")
```

```{r wilcoxon.cluster,cache=TRUE}
wilcoxon.cluster.gr <- adply(.data=cluster.data[, -1], 2, 
                             .fun=do.wilcoxon,
                             classif=cluster.data[, 1], 
                             alternative="g")

colnames(wilcoxon.cluster.gr) <- c("Species", "p.val")

wilcoxon.cluster.gr <- transform(wilcoxon.cluster.gr, 
                                 p.adj=p.adjust(p.val, method="BH"))
wilcoxon.cluster.gr <- wilcoxon.cluster.gr[order(wilcoxon.cluster.gr$p.adj, 
                                                decreasing=FALSE), ]
wilcoxon.cluster.gr <- subset(wilcoxon.cluster.gr, p.adj <= 0.05)

wilcoxon.cluster.le <- adply(.data=cluster.data[, -1], 2,
                             .fun=do.wilcoxon,
                             classif=cluster.data[, 1],
                             alternative="l")

colnames(wilcoxon.cluster.le) <- c("Species", "p.val")
wilcoxon.cluster.le <- transform(wilcoxon.cluster.le, 
                                 p.adj=p.adjust(p.val, method="BH"))
wilcoxon.cluster.le <- wilcoxon.cluster.le[order(wilcoxon.cluster.le$p.adj, 
                                                 decreasing=FALSE), ]
wilcoxon.cluster.le <- subset(wilcoxon.cluster.le, p.adj <= 0.05)
```

```{r diff.cluster.gr,results='asis'}
tbl.cluster.gr <- xtable(wilcoxon.cluster.gr)
print(tbl.cluster.gr, type="html")
```

```{r diff.cluster.le,results='asis'}
tbl.cluster.le <- xtable(wilcoxon.cluster.le)
print(tbl.cluster.le, type="html")
```

A summary of *Lactobacillus gasseri_361* abundances, which is one that is cited as being significantly different between the T2D and NGT groups using the MGCs:

```{r boxplot1,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
p <- ggplot(cluster.table, aes(x=Classification, y=Lactobacillus.gasseri_361))
p <- p + geom_boxplot() + scale_y_log10()
p
```

A summary of *Lactobacillus gasseri JV-V03* abundances, which is one that is cited as being significantly different between the T2D and NGT groups using the species:

```{r boxplot2,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
p <- ggplot(strain.table, aes(x=Classification, y=Lactobacillus.gasseri.JV.V03))
p <- p + geom_boxplot() + scale_y_log10()
p
```

Let's try to train a random forest, first using the strain abundances. First, in figure 1 the authors say they used 915 species; however, their species/strain table has 2075 species. 

Now, for the random forest training:

```{r rfstrains, cache=TRUE}
rf.strains <- randomForest(Classification ~ ., data=strain.data,
                           mtry=500, replace=TRUE, ntree=10000,
                           importance=TRUE, proximity=TRUE,
                           do.trace=1000)
```

```{r importancestrains,results='asis'}
importance.strains <- rf.strains$importance
importance.strains <- importance.strains[order(importance.strains[, "MeanDecreaseAccuracy"], 
                                               decreasing=TRUE), ]
importance.table <- xtable(head(importance.strains, 30))
print(importance.table, type="html")
```

Let's look at how the data clusters using the proximities from the random forest.

```{r mdsstrains,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
MDSplot(rf.strains, strain.data$Classification)
```

Now, let's train a random forest using the MGCs.

```{r rfcluster, cache=TRUE}
rf.cluster <- randomForest(Classification ~ ., data=cluster.data,
                           mtry=500, replace=TRUE, ntree=10000,
                           importance=TRUE, proximity=TRUE,
                           do.trace=1000)
```

Here are the top 30 variables ranked by accuracy using the MGCs as features:
```{r importanceclusters,results='asis'}
importance.clusters <- rf.cluster$importance[order(rf.cluster$importance[, "MeanDecreaseAccuracy"], 
                                                   decreasing=TRUE), ]
importance.table <- xtable(head(importance.clusters, 30))
print(importance.table, type="html")
```


```{r mdsclusters,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
MDSplot(rf.cluster, cluster.data$Classification)
```