Gut metagenome in European women with normal, impaired and diabetic glucose control
========================================================

```{r loadlibs,echo=FALSE, warning=FALSE,message=FALSE}
library(randomForest)
library(ggplot2)
library(xtable)
```


This is an attempt to recreate the figures in the paper "Gut metagenome in European women with normal, impaired and diabetic glucose control". Their methods included the training of a random forest for diabetes prediction. The data used for training was obtained from the authors.


```{r readdata,cache=TRUE}
cluster.table <- read.delim("data/ClusterTable.txt", sep="\t", header=TRUE)
strain.table <- read.delim("data/StrainTable.txt", sep="\t", header=TRUE)
supptable3 <- read.csv("data/SuppTable3.csv")
```
They focus on the Type II diabetes (T2D) and normal glucose tolerance (NGT) classes. There is another class present in the sample (impaired glucose tolerance; IGT), but I will remove these data points.

```{r subsetsupptable,cache=TRUE}
supptable3 <- subset(supptable3, Classification != "IGT")
supptable3 <- transform(supptable3, Classification=factor(Classification))
```

We need to combine the classifications with the abundances:
```{r combinedata,cache=TRUE}
cluster.table <- merge(supptable3[, c("Sample.ID", "Classification")], cluster.table, by.x="Sample.ID", by.y="Subjects")
strain.table <- merge(supptable3[, c("Sample.ID", "Classification")], strain.table, by.x="Sample.ID", by.y="Subject")
```

A view of number of subjects per group:

```{r histogram,fig.width=7,fig.height=6,echo=TRUE}
p <- ggplot(cluster.table, aes(x=Classification))
p <- p + geom_bar()
p
```

A summary of *Lactobacillus gasseri_361* abundances, which is one that is cited as being significantly different between the T2D and NGT groups using the MGCs:

```{r boxplot1,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
p <- ggplot(cluster.table, aes(x=Classification, y=Lactobacillus.gasseri_361))
p <- p + geom_boxplot() + scale_y_log10()
p
```

A summary of *Lactobacillus gasseri JV-V03* abundances, which is one that is cited as being significantly different between the T2D and NGT groups using the MGCs:

```{r boxplot2,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
p <- ggplot(strain.table, aes(x=Classification, y=Lactobacillus.gasseri.JV.V03))
p <- p + geom_boxplot() + scale_y_log10()
p
```

Let's try to train a random forest, first using the strain abundances. First, in figure 1 the authors say they used 915 species; however, their species/strain table has 2075 species. I'm assuming they used the 915 most abundant ones, so let's compute that.

```{r topstrains,message=FALSE}
strain.data <- strain.table[, -1]
strain.mean.abundances <- apply(strain.data[, -1], 2, mean)

# Add one to account for the column I left out
strain.mean.abundances.ordered <- order(strain.mean.abundances, decreasing=TRUE) + 1

# Top 915 by mean
strain.data.top <- strain.data[, c(1, head(strain.mean.abundances.ordered, 915))]

```

Now, for the random forest training:

```{r rfstrains, cache=TRUE}
rf.strains <- randomForest(Classification ~ ., data=strain.data.top,
                           mtry=500, replace=TRUE, ntree=10000,
                           importance=TRUE, proximity=TRUE,
                           do.trace=1000)
```

```{r importancestrains,results='asis'}
importance.strains <- rf.strains$importance
importance.strains <- importance.strains[order(importance.strains[, "MeanDecreaseAccuracy"], decreasing=TRUE), ]
importance.table <- xtable(head(importance.strains, 30))
print(importance.table, type="html")
```

Let's look at how the data clusters using the proximities from the random forest.

```{r mdsstrains,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
MDSplot(rf.strains, strain.data.top$Classification)
```

Now, let's train a random forest using the MGCs.

```{r rfcluster, cache=TRUE}
cluster.data <- cluster.table[, -1]
rf.cluster <- randomForest(Classification ~ ., data=cluster.data,
                           mtry=500, replace=TRUE, ntree=10000,
                           importance=TRUE, proximity=TRUE,
                           do.trace=1000)
```

Here are the top 30 variables ranked by accuracy using the MGCs as features:
```{r importanceclusters,results='asis'}
importance.clusters <- rf.cluster$importance[order(rf.cluster$importance[, "MeanDecreaseAccuracy"], decreasing=TRUE), ]
importance.table <- xtable(head(importance.clusters, 30))
print(importance.table, type="html")
```


```{r mdsclusters,fig.width=7, fig.height=6,echo=FALSE,warning=FALSE}
MDSplot(rf.cluster, cluster.data$Classification)
```